biological data preperation

### 2018-09-06 src/biological_data_prep.Rmd
back up before R restart for package installation

# Overview
Preperation of the capelin (*Mallotus villosus*) data obtained from [OBIS](www.iobis.org). This database shows the global occurance of capelin. 
Built with 'r getRversion()'.

# Package dependencies
You can load them using the following code which uses a function called [ipak](https://gist.github.com/stevenworthington/3178163). 
Note this function checks to see if the packages are installed first.
Th
```{r pre-install packages, message=FALSE}
packages <- c("packrat", "rworldmap", "dplyr" "plyr")
source("../src/ipak.R")
ipak(packages)
suppressMessages(ipak)
```

@@

```{r}
library("dplyr")
tidydata <- distinct(tidydata, decimalLatitude, decimalLongitude, datecollected, institutioncode, institutioncode, individualcount, resname, originalscientificname)
```

@@

Looking at the maps, the ROM, NAFO, IMR, ICES have data next to iceland.
Since NAFO, IMR, and ICES only have one data point each, you can remove these observations easily from the record...
```{r}
tidydata <- tidydata[!tidydata$institutioncode == "NAFO", ]
tidydata <- tidydata[!tidydata$institutioncode == "IMR", ]
tidydata <- tidydata[!tidydata$institutioncode == "ICES", ]
```
Now for  ROM there are more data points...

write what you have so far to a csv
```{r}
write.csv(tidydata, file = "../data/bio/capelin_obis_tidy.csv", row.names = FALSE)
```

>>>>

# Overview
Preperation of the capelin (*Mallotus villosus*) data obtained from [OBIS](www.iobis.org). This database shows the global occurance of capelin.

A note to anyone who might happen to stumble across this... I am a beginner in R and have had no exposure to similar languages. The code herein is unlikely to be elegant and there is likely more efficient ways of running the code.

Built with 'r getRversion()'.

# Package dependencies
You can load them using the following code which uses a function called [ipak](https://gist.github.com/stevenworthington/3178163). 
Note this function checks to see if the packages are installed first.
Th
```{r pre-install packages, message=FALSE}
packages <- c("packrat", "rworldmap", "dplyr","plyr", "rgdal")
source("../src/ipak.R")
ipak(packages)
suppressMessages(ipak)
```

@@

```{r}
library("dplyr")
tidydata <- distinct(tidydata, decimalLatitude, decimalLongitude, datecollected, institutioncode, institutioncode, individualcount, resname, originalscientificname, .keep_all = T) #the .keep_all means that id isn't excluded from the output
head(tidydata)
```

@@

Looking at the maps, the ROM, NAFO, IMR, ICES have data next to iceland.
Since NAFO, IMR, and ICES only have one data point each, you can remove these observations easily from the record...
```{r remove iceland data}
tidydata <- tidydata[!tidydata$institutioncode == "NAFO", ]
tidydata <- tidydata[!tidydata$institutioncode == "IMR", ]
tidydata <- tidydata[!tidydata$institutioncode == "ICES", ]
```

Now for  ROM there are more data points...
Use indentify function which lets you click on a point on the plot
```{r locate coordinates of iceland point}
x11() #this opens the plot i a window you can interact with
plot(map2, xlim = c(-71, -43), ylim =c(39, 62), asp = 1, main = "Royal Ontario Museum Occurrences",  col = "cornsilk") #the x and y lim are the long-lat bounds of the map
points(ROM_obs$decimalLongitude, ROM_obs$decimalLatitude, col = "red") #this adds points to the mapet", xlab = "Longitude", ylab = "Latitude"))
identify(ROM_obs$decimalLongitude, ROM_obs$decimalLatitude, labels=ROM_obs$decimalLongitude) #decimalLongitude chosen as plot identifier as the values are unique
```
The point that needs to be removed is located at -44.96667. Note this is not the full value so use some other locators to remove the row (you could just used id, but pick multiple others just to learn)

```{r remove iceland point}
tidydata <- tidydata[!(tidydata$resname == "Fish specimens" & tidydata$datecollected == "1965-08-13 12:00:00+01"), ]
```

now replot all data just to check...

```{r map occurrences less iceland}
plot(map2, xlim = c(-71, -43), ylim =c(38, 62), asp = 1, main = "All Occurrences without Iceland", col = "cornsilk") #the x and y lim are the long-lat bounds of the map
points(tidydata$decimalLongitude, tidydata$decimalLatitude, col = "red") #this adds points to the mapet", xlab = "Longitude", ylab = "Latitude")
dev.copy(png, "../output/bio/all_occurrences_less_iceland.png") #this prints a png of the plot
dev.off() #this turns off the print command
```
great! write what you have so far to a csv
```{r}
write.csv(tidydata, file = "../data/bio/capelin_obis_tidy.csv", row.names = FALSE)
```

# Remove dates you don't have environmental data for
The BIOMER dataset starts in 1998-01 (GLORYS 1993), so cut all occurrences prior to 1998-01

```{r remove occurrences prior to 98/01}
tidytest <- tidydata
tidytest <- tidytest[tidytest$year >= 1998, ]
```

and write to a csv...
```{r}
write.csv(tidydata, file = "../data/bio/capelin_obis_tidy.csv", row.names = FALSE)
```

# Adding new (non-environmental) data to the file
There are some more columns you want to add...
- Occurrence - presence/background (1/0)
- NAFO region
- Gear type
- Unique Cell ID (based on the environmental data)

so not to mess with tidydata, create a new variable and write a new .csv
```{r create new variable/csv for adding data}
adddata <- tidydata
write.csv(adddata, file = "../data/bio/capelin_additional_tidy.csv", row.names = FALSE)
```

first create a new column for Occurence, and populate with "1" (because these are all presences)
```{r}
adddata$occurrence <- "1"
head(adddata)
```

# Add NAFO Region occurrence is from
NAFO Zones shapefile obtained from [NAFO](https://www.nafo.int/Data/GIS)

```{r}
library(rgdal)
addtest <- adddata

```
